---
title: "R Class #2"
author: "Marcel Roman"
date: "10/30/2018"
output: html_document
---

# Taking a step back --- learning about different types of vectors 

There are different types of vectors/data/variables, and these types possess different properties that matter for inference and estimation. 

 * **Character Vectors/Variables:** These are variables/vectors that are like characters. For instance, a character vector may have five different elements, "hello", "my," "name," "is," "marcel," which could be expressed as \{"hello", "my", "name", "is", "marcel"\}. When functions are run over character vectors, they are typically transformed into factor categories, where the estimates are comparisons between different categories. Think of the classic race variable, which could be expressed as a character variable, \{"black", "white", "asian", "latinx", "other"\}. 

 * **Factor Vector/Variables:** These may look like character variables/vectors in $\mathbb{R}$, but they are considered "categories", that are latently expressed as numbers but the order does not have any bearing on the meaning of the way the variable is measured itself. The details of how this works will be hammered out when we work with this type of data. 

 * **Numeric Vector/Variables:** These are just numbers, e.g. \{1, 2, 3, 4, 5\} is a 5 element vector containing the elements 1, 2, 3, 4, and 5. 
 
Let's break these down with some code. 

### Character Vectors 

```{r}

# install.packages('stringr')

library(stringr)

# Character vectors 

char_vect1 = c('black', 'latinx', 'white', 'asian')
char_vect2 = c('male', 'female', 'trans', 'non-binary')

# You can't run certain functions over character vectors like 
# you could with numeric vectors. For example, if we decided
# to take the average of a character vector, you receive an 
# error

mean(char_vect1)

# But there are other functions that are made specifically 
# for manipulating character vectors, for instance, the 
# paste() function 

output_char1 = paste(char_vect1, char_vect2, sep = '/')
output_char1

output_char2 = paste(char_vect1, char_vect2, collapse = '|')
output_char2

# How many elements does output_char1 have? How many elements 
# does output_char2 have? 

length(output_char1)
length(output_char2)

# There are also some interesting manipulations you can run 
# using base R and the stringr package

toupper(char_vect1)
toupper(char_vect2)

tolower(toupper(char_vect1))
tolower(toupper(char_vect2))

str_to_upper(char_vect1)
str_to_upper(char_vect2)

# Say you want a 12 character element for all elements involved padded
# with zeros (leading zeros)

str_pad(string = char_vect1, width = 12, side = 'left', pad = '0')
str_pad(string = char_vect2, width = 12, side = 'left', pad = '0')

# Say you want to substring to the first three characters of the 
# elements

substring(char_vect1, 1, 3)
substring(char_vect2, 1, 3)

# Okay, now what if we want to choose elements from the vector
# on the basis of a string pattern? Use the grep function, 
# which is INCREDIBLY HANDY! 

  # Value version --- outputs character

grep(x = char_vect1, pattern = 'black', value = TRUE)

  # Index version --- outputs numeric index

grep(x = char_vect1, pattern = 'black', value = FALSE)
char_vect1[grep(x = char_vect1, pattern = 'black', value = FALSE)]

# What if we want to replace parts of a character vector based on 
# some sort of pattern? 

char_vect1_a = paste(char_vect1, 'person')

# Let's replace person with "individual"

gsub(x = char_vect1_a, pattern = 'person',
     replacement = 'individual')

# Also useful for removing whitespace

gsub(x = char_vect1_a, pattern = ' ', replacement = '')


# For more on character manipulation, see 

# https://www.gastonsanchez.com/Handling_and_Processing_Strings_in_R.pdf


```

### Factor Vectors 

```{r}

# Factor variables are a little weird, so let's try to construct 
# one from scratch. 

fact_vect = c(1, 2, 3, 4, 5)
fact_vect = factor(fact_vect, labels = c('white', 'black', 'latinx', 
                             'asian', 'other'))

# What happens when you convert a factor?

as.numeric(fact_vect)
as.numeric(as.character(fact_vect))

# You can exclude pieces of a factor variable 

factor(fact_vect, exclude = 'latinx')

# Remember, factors have reference categories, which you can check 
# with the levels() function --- the first level is the reference

levels(fact_vect)

# You can "relevel" the reference category

relevel(x = fact_vect, ref = 'latinx')

```

### Numeric Vectors 

```{r}

# Numeric Vectors 

num_vect1 = c(1, 2, 3, 4, 5)
num_vect2 = c(11, 12, 13, 14, 15)

# We can run functions on numeric vectors

mean(num_vect1)
mean(num_vect2)

log(num_vect1)
log(num_vect2)

# If numeric vectors have the same number of elements, we can pass 
# mathematical operations on them such as addition, subtraction, 
# division, multiplication, etc. 

num_vect1 + num_vect2

num_vect1 - num_vect2

num_vect1 / num_vect2

num_vect1 * num_vect2

```

Mathematical operations with numeric vectors are conducted *element-wise* --- that is, the first element in the first vector, $X_{i}$ is multiplied/added to/subtracted from/divided by the first element in the second vector, $Y_{i}$, and so on and so forth. 

# Matrices and dataframes

Datasets in R are formulated in an $n \times k$ format, just like matrices. They are constructed of rows and columns. But there are differences between *matrices*, and *dataframes*, the subtleties and distinctions we will discuss. 


```{r}

library(tidyverse)
library(dplyr)

# Let's construct a matrix from scratch

set.seed(2317)

norm_vect = rnorm(10)
race_vect = c('black', 'black', 'white', 'asian', 'asian',
              'asian', 'latinx', 'black', 'white', 'black')
norm_vect2 = rnorm(10, 3)

mat = cbind(norm_vect, norm_vect2, race_vect)

# Now let's analyze the properties of this matrix

  # Weirdness because the cbind() function made our numeric vectors
  # into characters

str(mat)
glimpse(mat)

# You could also generate a matrix via the matrix() function, but 
# it doesn't fix the character problem 

matrix(c(norm_vect, norm_vect2, race_vect), nrow = 10, ncol = 3)

# You can select rows and columns or elements within rows and 
# columns on the index using []

mat[1, ]
mat[, 1]
mat[, 3]

mat[1, 1]
mat[1, 3]

# Okay, so let's try to fix the fact that our matrix has some 
# character vectors 

mat[, 1] = as.numeric(mat[, 1])
mat[, 2] = as.numeric(mat[, 2])

mat = cbind(norm_vect, norm_vect2, race_vect)
mat[, 1:2] = apply(X = mat[, 1:2], 2, function(x) as.numeric(x))

# Problem: Any non-numeric vector will force the matrix to be character

mat2 = cbind(apply(X = mat[, 1:2], 2, function(x) as.numeric(x)),
      norm_vect)

# We can apply functions over columns 

mean(mat2[, 1])
mean(mat2[, 2])
mean(mat2[, 3])

# We can apply functions over rows

mean(mat2[1, ])

# Matrix properties apply --- addition 

mat3 = cbind(rnorm(10, 5), rnorm(10, 3, 2), rnorm(8, 2, 1))

mat2 + mat3

# Subtraction

mat2 - mat3

# Multiplication (element wise)

mat2 * mat3

# Division (which will occur element wise)

mat2 / mat3

# Multiplication (matrix algebra)

# mat2 %*% mat3

mat2 %*% t(mat3)


# Let's convert our original matrix into a dataframe

df1 = as.data.frame(cbind(norm_vect, norm_vect2, race_vect))

# DFs are nice because we can call column-wise vectors easily 

df1$norm_vect
df1$norm_vect2
df1$race_vect

# We can also convert select variables without problems 

df1$norm_vect = as.numeric(as.character(df1$norm_vect))
df1$norm_vect2 = as.numeric(as.character(df1$norm_vect2))
df1$race_vect = as.character(df1$race_vect)

str(df1)
glimpse(df1)


# These kind of behave in similar ways as matrices, 
# but they're not matrices

df1$norm_vect * df1$norm_vect2
df1$norm_vect / df1$norm_vect2
df1$norm_vect + df1$norm_vect2
df1$norm_vect - df1$norm_vect2

# df1$race_vect * df1$norm_vect2

# You can run several manipulations on dataframes 

colnames(df1) = c('var1', 'var2', 'race')

# You can generate new vectors quite easily 

df1$var3 = NA
df1$var3 = df1$var1 * 3

df1$var4 = NA
df1$var4 = df1$var1 / df1$var2

# An aside, you can call by variable name 

df1[, 'var1']
df1[, 'race']

# You can also use various dplyr functions on dataframes 

df1_a = df1 %>% mutate(var5 = var2 * var3) %>% 
  rename(first_var = var1) %>% 
  mutate(race = recode(race, `black` = 'b', `white` = 'w',
                       `asian` = 'a', `latinx` = 'l'))

df1_b = df1_a %>% filter(race == 'b')

# Think of dplyr as running functions through a filter, or 
# "pipe", %>% 


```

### Lists 

Lists are all-purpose $\mathbb{R}$ objects which can contain elements of almost anything: numbers, strings, vectors, another list inside of it, dataframes and matrices. So they're quite flexible and useful for storing various $\mathbb{R}$ objects you may want to use for whichever purpose. 

```{r}

# Let's create a list 

list_data = list('One', 'Two', c(30, 25, 25), TRUE, 51.3281, 88.1354)
list_data

# Lists are indexed a little differently 

list_data[[3]]

# You can put an entire dataframe in a list 

list_data[[6]] = df1
list_data[[6]]

# Elements within elements in a list 

list_data[[3]][3]

list_data[[6]]$var1

list_data[6]

# Resource, see here for more on lists:
# https://www.tutorialspoint.com/r/r_lists.htm

```

# Functions 

You can generate functions in $\mathbf{R}$. Remember the structure of functions, they are made up of parameters. These parameters will be feed into a process that constitutes the function. Let's see how this works in practice. 

```{r}

# Let's say we wanted a simple function that takes a scalar object and # adds two to it 

x = 2 

add_2 = function(x) {
  
  x + 2
  
}

add_2(x)

# Lets change the input 

newinput = 312381
add_2(newinput)

# Now let's see if a vector will work? 

newinput2 = c(1, 2, 3, 4, 5)
add_2(newinput2)

# Functions can be complicated. Let's say we wanted a function that will give us TRUE if the number is negative, FALSE if the number is positive. Then we want to employ an if else condition. 

newf2 = function(x) { 
  
  if (x >= 0) {
    
    print(TRUE)
    
  } else {
    
    print(FALSE)
    
  }
  
}

x = 2
y = 0
z = -5

newf2(x)
newf2(y)
newf2(z)

```


# Loops 

You can also generate loops. This allows you to iterate a process over a series of indices to serve some purpose. For instance, what if I wanted to get the means of the columns of a dataframe and store those means into a single vector? Here's where loops come into play: 


```{r} 

set.seed(3128731)

dat = as.data.frame(cbind(
  rnorm(1000), rnorm(1000), rnorm(1000), rnorm(1000), rnorm(1000),
  rnorm(1000), rnorm(1000), rnorm(1000), rnorm(1000), rnorm(1000)
))

# Theoretically I could take the average of each row and plug it into
# a vector

vect_storage = rep(NA, 10)
vect_storage[1] = mean(dat$V1)
vect_storage[2] = mean(dat$V2)
vect_storage[3] = mean(dat$V3)
vect_storage[4] = mean(dat$V4)
vect_storage[5] = mean(dat$V5)
vect_storage[6] = mean(dat$V6)
vect_storage[7] = mean(dat$V7)
vect_storage[8] = mean(dat$V8)
vect_storage[9] = mean(dat$V9)
vect_storage[10] = mean(dat$V10)

# But this is alot of code. Let's reduce the amount of code we need

reps = 10 
vect_storage = rep(NA, 10)
for (i in 1:reps) {
  
  vect_storage[i] = mean(dat[, i])
  
}

# You could also use the apply function

vect_storage = as.numeric(apply(X = dat, MARGIN = 2, FUN = mean))
vect_storage

# There are various apply functions that work for lists and vectors
# e.g lapply() and vapply()


```

# First Cut at working with real data


You need to set your working directory first: 

* Save your script to a directory and set the working directory (Top Bar: Session > To Source File Location)

* Generate a project directory, and work from there without having to set the working directory. 


```{r} 

library(foreign)
library(haven)

# Importing the data --- it is a .dta extension file, so we need 
# to use the appropriate read function

wt_dta = read_stata('weight_data.dta')

wt_dta = as.data.frame(wt_dta)

# Looking at the data 

glimpse(wt_dta)

# We can do some simple analysis 

mean(wt_dta$wt[wt_dta$hair == 'BLACK'])
mean(wt_dta$wt[wt_dta$hair == 'BLONDE'])
mean(wt_dta$wt[wt_dta$hair == 'BROWN'])

# 

mean(wt_dta$shoe[wt_dta$hair == 'BLACK'])
mean(wt_dta$shoe[wt_dta$hair == 'BLONDE'])
mean(wt_dta$shoe[wt_dta$hair == 'BROWN'])

# Blonde vs. everyone else

mean(wt_dta$shoe[wt_dta$hair == 'BLONDE'])
mean(wt_dta$shoe[wt_dta$hair == 'BLACK' | wt_dta$hair == 'BROWN'])

# We can generate new variables

wt_dta$shoe_wt = wt_dta$shoe / wt_dta$wt
mean(wt_dta$shoe_wt)
wt_dta = wt_dta %>% mutate(shoe_wt = shoe / wt)


```


### Assignment


```{r}

# 1) Generate a new dataset that is the same as the weight/shoe size 
# data but only includes people with blonde hair (show that you can 
# subset the data in two ways: a) Using the indices, b) Using the 
# filter() command in dplyr with the %>% operator) 





# 2) Generate a new variable that is weight and shoe size multiplied
# by one another (use the dplyr %>% operator and mutate() function for 
# this problem)






# 3) Take the mean of the new variable for people with black hair 





# 4) Take the mean of the new variable for people with blonde hair 





# 5) Do 3 and 4 but derive the median instead 





# 6) Take the original data, group by hair color, and generate a new "by-group"
# dataset with the mean shoe size and weight of each hair color group. Use 
# the dplyr %>% operator, group_by(), and summarize() functions. 





```