---
title: "R Class - Session 1"
author: "Ana Oaxaca"
date: "10/14/2018"
output: html_document
---

# Downloading + Installing

### R
R is an open source statistical language, to make sense of data through computing and graphics. R provides a wide variety of statistical (linear and nonlinear modelling, classical statistical tests, time-series analysis, classification, clustering, …) and graphical techniques, and is highly extensible. 

http://cran.stat.ucla.edu

### R Studio
RStudio is a free and open-source integrated development environment (IDE) for R, a programming language for statistical computing and graphics. Basically an editor for R.

https://www.rstudio.com/products/rstudio/download/

### Latex
LaTeX is a high-quality typesetting system; it includes features designed for the production of technical and scientific documentation. LaTeX is the de facto standard for the communication and publication of scientific documents. Think about it as a word processor.

https://www.latex-project.org/get/

### Texmaker + Texshop
Texmaker is a cross-platform LaTeX editor that integrates many tools needed to develop documents with LaTeX, in just one application. 
http://www.xm1math.net/texmaker/

https://pages.uoregon.edu/koch/texshop/obtaining.html

# Intro to R 

### Quick Tour

In R Studio, you should see:

1. Upper-left: Program editor window, where you’ll write code.
2. Lower-left: Console, where you’ll submit commands to be executed.
3. Upper-right: Workspace lists variables in memory, History lists the commands you’ve previously submitted in the console.
4. Lower-right: Files in current working directory, Plots you’ve produced, Packages installed and whether they’re loaded, Help on commands.

Keyboard Shortcuts: 

* Ctrl-Enter (submit current line or selection from editor window to console) Ctrl-1 (move cursor to editor window)*
* Ctrl-2 (move cursor to console)*
* Ctrl-6 (move cursor to plots)*
* Esc (interrupt currently executing command) Ctrl-L (clear console)*

An exhaustive list:

https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts

**Workflow** is how you get things done. Here’s my recommendation:

1. Work out commands in the editor, then copy/paste into console or better use Ctrl-Enter to submit current line or selection.
2. Your editor will keep a history of what you’ve done.
3. Make comments to help your future-self recall why you did what you did.

### R Building Blocks
```{r setup, echo=F}
knitr::opts_chunk$set(warning = F)
knitr::opts_chunk$set(message = F)
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(eval = F)

```

#### R as a calculator 
In the following sections, look at the commands and output
and understand how the command was interpretted to give the output.

```{r}
#### Calculator # Arithmetic
2 * 10
## [1] 20

1 + 2
## [1] 3

# Order of operations is preserved
1 + 5 * 10

## [1] 51
(1 + 5) * 10
## [1] 60

# Exponents use the ^ symbol
2^5
## [1] 32

9^(1/2)
## [1] 3
```

#### Vectors
A vector is a set of numbers like a column of a spreadsheet. Below these sets of numbers are not technically “vectors”, since they are not in a row/column structure, though they are ordered and can be referred to by index.

```{r}
#### Vectors
# Create a vector with the c (short for combine) function 
c(1, 4, 6, 7)
## [1] 1 4 6 7

c(1:5, 10)
## [1]  1  2  3  4  5 10

# or use a function
# (seq is short for sequence) 
seq(1, 10, by = 2)
## [1] 1 3 5 7 9

seq(0, 50, length = 11)
## [1] 0 5101520253035404550
seq(1, 50, length = 11)
## [1] 1.0 5.9 10.8 15.7 20.6 25.5 30.4 35.3 40.2 45.1 50.0 1:10 
# short for 

seq(1, 10, by = 1) # or just
## [1] 1 2 3 4 5 6 7 8 9 10

seq(1, 10)
## [1] 1 2 3 4 5 6 7 8 9 10

5:1
## [1] 5 4 3 2 1

# non-integer sequences
# (Note: the [1] at the beginning of lines indicates # the index of the first value in that row) 
seq(0, 100*pi, by = pi)
##   [1]   0.000000   3.141593   6.283185   9.424778  12.566371
##   [6]  15.707963  18.849556  21.991149  25.132741  28.274334
##  [11]  31.415927  34.557519  37.699112  40.840704  43.982297
##  [16]  47.123890  50.265482  53.407075  56.548668  59.690260
##  [21]  62.831853  65.973446  69.115038  72.256631  75.398224
##  [26]  78.539816  81.681409  84.823002  87.964594  91.106187
##  [31]  94.247780  97.389372 100.530965 103.672558 106.814150
##  [36] 109.955743 113.097336 116.238928 119.380521 122.522113
##  [41] 125.663706 128.805299 131.946891 135.088484 138.230077
##  [46] 141.371669 144.513262 147.654855 150.796447 153.938040
##  [51] 157.079633 160.221225 163.362818 166.504411 169.646003
##  [56] 172.787596 175.929189 179.070781 182.212374 185.353967
##  [61] 188.495559 191.637152 194.778745 197.920337 201.061930
##  [66] 204.203522 207.345115 210.486708 213.628300 216.769893
##  [71] 219.911486 223.053078 226.194671 229.336264 232.477856
##  [76] 235.619449 238.761042 241.902634 245.044227 248.185820
##  [81] 251.327412 254.469005 257.610598 260.752190 263.893783
##  [86] 267.035376 270.176968 273.318561 276.460154 279.601746
##  [91] 282.743339 285.884931 289.026524 292.168117 295.309709
##  [96] 298.451302 301.592895 304.734487 307.876080 311.017673
## [101] 314.159265
```

#### Variables 
Assigning objects to varibles.

```{r}
#### Assign variables
# Assign a vector to a variable with <- 
a <- 1:5
a
## [1] 1 2 3 4 5

b <- seq(15, 3, length = 5)
b
##[1]15 12 9 6 3

c <- a * b 
c
## [1] 15 24 27 24 15

```


#### Basic functions
There are lots of functions available in the base package. Type ?base and click on Index at the bottom of the help page for a complete list of functions in base R. 

```{r}
#### Basic functions
# Lots of familiar functions work 
a
## [1] 1 2 3 4 5

sum(a)
## [1] 15

prod(a)
## [1] 120

mean(a)
## [1] 3

sd(a)
## [1] 1.581139

var(a)
## [1] 2.5

min(a)
## [1] 1

median(a)
## [1] 3

max(a)
## [1] 5

range(a)
## [1] 1 5
```

#### Extracting Subsets
It will be an extremely important skill to understand the structure of your variables and pull out the information you need from them.

```{r}
#### Extracting subsets
# Specify the indices you want in the square brackets []
a <- seq(0, 100, by = 10)
a
##[1] 0 10 20 30 40 50 60 70 80 90 100

a[] # blank = include all
##[1] 0 10 20 30 40 50 60 70 80 90 100

# integer + = include, 0 = include none, - = exclude
a[5]
## [1] 40

a[c(2, 4, 6, 8)]
## [1] 10 30 50 70

a[0]
## numeric(0)

a[-c(2, 4, 6, 8)]
##[1] 0 20 40 60 80 90 100

a[c(1, 1, 1, 6, 6, 9)] # subsets can be bigger than the original set
##[1] 0 0 0 50 50 80

a[c(1,2)] <- c(333, 555) # update a subset
a
## [1]333 555 20 30 40 50 60 70 80 90 100
```


#### Boolean
True/False Subsets can be selected based on which elements meet specific conditions.

```{r}
#### Boolean
a
## [1]333 555 20 30 40 50 60 70 80 90 100

(a > 50) # TRUE/FALSE for each element
## [1] TRUE TRUE FALSE FALSE FALSE FALSE TRUE TRUE TRUE TRUE TRUE 

which(a > 50) # which indicies are TRUE
## [1] 1 2 7 8 9 10 11

a[(a > 50)]
##[1]333555 60 70 80 90 100

!(a > 50) # ! negates (flips) TRUE/FALSE values
## [1] FALSE FALSE TRUE TRUE TRUE TRUE FALSE FALSE FALSE FALSE FALSE 

a[!(a > 50)]
## [1] 20 30 40 50
```

### Comparison functions
All your favorite comparisons are available.
```{r}
#### Comparison
#Here they are: < > <= >= != == %in% 
a
## [1]333555 20 30 40 50 60 70 80 90 100

# equal to
a[(a == 50)]
## [1] 50

# equal to
a[(a == 55)]

## numeric(0)
# not equal to
a[(a != 50)]
## [1]333555 20 30 40 60 70 80 90100

# greater than
a[(a > 50)]
##[1]333555 60 70 80 90 100

# less than
a[(a < 50)]
## [1] 20 30 40

# less than or equal to
a[(a <= 50)]
## [1] 20 30 40 50

  # which values on left are in the vector on right
(c(10, 14, 40, 60, 99) %in% a)
## [1] FALSE FALSE  TRUE  TRUE FALSE
```

#### Boolean operators 
Compare TRUE/FALSE values and return TRUE/- FALSE values.
```{r}
#### Boolean
# & and, | or, ! not 
a
## [1]333 555 20 30 40 50 60 70 80 90 100 

a[(a >= 50) & (a <= 90)]
## [1] 50 60 70 80 90

a[(a < 50) | (a > 100)]
##[1]333555 20 30 40

a[(a < 50) | !(a > 100)]
## [1] 20 30 40 50 60 70 80 90 100 

a[(a >= 50) & !(a <= 90)]
## [1] 333 555 100
```

#### Missing values 
The value NA (not available) means the value is missing. Any calculation involving NA will return an NA by default.

```{r}
#### Missing values
NA + 8
## [1] NA

3 * NA
## [1] NA

mean(c(1, 2, NA))
## [1] NA

# Many functions have an na.rm argument  (NA remove)
mean(c(NA, 1, 2), na.rm = TRUE)
## [1] 1.5

sum(c(NA, 1, 2))
## [1] NA

sum(c(NA, 1, 2), na.rm = TRUE)
## [1] 3

# Or you can remove them yourself
a <- c(NA, 1:5, NA)
a
## [1] NA 1 2 3 4 5 NA

is.na(a) # which values are missing?
## [1] TRUE FALSE FALSE FALSE FALSE FALSE TRUE

!is.na(a) # which values are NOT missing?
## [1] FALSE TRUE TRUE TRUE TRUE TRUE FALSE 

a[!is.na(a)] # return those which are NOT missing
## [1] 1 2 3 4 5

a # note, this did not change the variable 
## [1] NA 1 2 3 4 5 NA

# To save the results of removing the NAs,
# assign to another variable or reassign to the original variable
# Warning: if you write over variable a then the original version is gone forever!
a <- a[!is.na(a)]
a
## [1] 1 2 3 4 5
```

 
### R command review
This is a summary of the commands we’ve seen so far.
```{r}
#### Review
<-
+-*/^
c()
seq() # by=, length=
sum(), prod(), mean(), sd(), var(), min(), median(), max(), range() a[]
(a > 1), ==, !=, >, <, >=, <=, %in% &, |, !
NA, mean(a, na.rm = TRUE), !is.na()

````